Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-07-02 submission/abstract_file.tex
minor edits

In many practical applications, one works with deep neural network (DNN) models trained by someone else.
For such \emph{pretrained models}, one often does not have access to training data or test data.
Moreover, one may not know many details about the model, such as the specifics of the training data, the loss function, the hyperparameter values, etc.
Given one or many pretrained models, can one say anything about the expected performance or quality of the models?
Here, we provide a detailed empirical meta-analysis to address this question, evaluating quality metrics for pretrained DNN models at scale.
%%Our metrics are drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as Heavy-Tailed Random Matrix Theory (HT-RMT), as used in the recently-developed Theory of Heavy-Tailed Self Regularization (e.g., fitted power law metrics used to characterize the degree of strong correlations in trained models).
%%The most promising are metrics drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as metrics (e.g., fitted power law metrics used to characterize the degree of strong correlations in a system) derived from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
%The most promising are norm-based metrics
% (such as those used to provide capacity control in traditional statistical learning theory) 
%and metrics based on fitting eigenvalue distributions to truncated power law (PL) distributions 
%(which derive from statistical mechanics, in particular the recently-developed Theory of Heavy-Tailed Self Regularization, and which characterize the strength of correlations in a DNN). 
%
Using the open-source \emph{WeightWatcher} tool to analyze DNN weight matrices, we analyze hundreds of publicly-available pretrained models, including older and current state-of-the-art models in computer vision (CV) and natural language processing (NLP).
We examine both familiar norm-based capacity control metrics (Frobenius and Spectral norms) as well as newer Power Law (PL) based metrics (including fitted PL exponents, $\alpha$, and the Weighted Alpha metric, $\hat{\alpha}$), from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
We also introduce the $\alpha$-Shatten Norm metric.
We find that norm-based metrics correlate well with reported test accuracies for well-trained models across nearly all CV architecture series.
On the other hand, we find that norm-based metrics can not distinguish well-trained versus poorly-trained models---which, arguably is the point of needing quality metrics.  
%%%Indeed, they may give spurious results.
We also find that PL-based metrics do much better---qualitatively better at discriminating well-trained versus poorly-trained models, and quantitatively better at discriminating among a series of well-trained versus more-well-trained models.
PL-based metrics can also be used to characterize fine-scale properties of these models.
%%%, and we introduce the layer-wise \emph{Correlation Flow} as new quality assessment.
%%%We show how poorly-trained (and/or poorly fine-tuned) models may exhibit both \emph{Scale Collapse} and unusually large PL exponents, $\alpha\gg 6$, in particular for recent NLP models.
Our techniques, as implemented in the \emph{WeightWatcher} tool, can be used to identify when a pretrained DNN has problems that can not be detected simply by examining training/test accuracies.

>>>>

In many practical applications, one works with deep neural network (DNN) models trained by someone else.
For such \emph{pretrained models}, one often does not have access to training data or test data.
Moreover, one may not know many details about the model, such as the specifics of the training data, the loss function, the hyperparameter values, etc.
Given one or many pretrained models, can one say anything about the expected performance or quality of the models?
Here, we provide a detailed empirical meta-analysis to address this question, evaluating quality metrics for pretrained DNN models at scale.
%%Our metrics are drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as Heavy-Tailed Random Matrix Theory (HT-RMT), as used in the recently-developed Theory of Heavy-Tailed Self Regularization (e.g., fitted power law metrics used to characterize the degree of strong correlations in trained models).
%%The most promising are metrics drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as metrics (e.g., fitted power law metrics used to characterize the degree of strong correlations in a system) derived from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
%The most promising are norm-based metrics
% (such as those used to provide capacity control in traditional statistical learning theory) 
%and metrics based on fitting eigenvalue distributions to truncated power law (PL) distributions 
%(which derive from statistical mechanics, in particular the recently-developed Theory of Heavy-Tailed Self Regularization, and which characterize the strength of correlations in a DNN). 
%
Using the open-source \emph{WeightWatcher} tool to analyze DNN weight matrices, we analyze hundreds of publicly-available pretrained models, including older and current state-of-the-art models in computer vision (CV) and natural language processing (NLP).
We examine both familiar norm-based capacity control metrics (Frobenius and Spectral norms) as well as newer Power Law (PL) based metrics (including fitted PL exponents, $\alpha$, and the Weighted Alpha metric, $\hat{\alpha}$), from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
We also introduce the $\alpha$-Shatten Norm metric.
We find that norm-based metrics correlate well with reported test accuracies for well-trained models across nearly all CV architecture series.
On the other hand, we find that norm-based metrics can not distinguish well-trained versus poorly-trained models---which, arguably is the point of needing quality metrics.  
%%%Indeed, they may give spurious results.
We also find that PL-based metrics do much better---qualitatively better at discriminating well-trained versus poorly-trained models, and quantitatively better at discriminating among a series of well-trained models \nred{of increasing depth and accuracy}.
%versus more-well-trained models.
PL-based metrics can also be used to characterize fine-scale properties of these models.
%%%, and we introduce the layer-wise \emph{Correlation Flow} as new quality assessment.
%%%We show how poorly-trained (and/or poorly fine-tuned) models may exhibit both \emph{Scale Collapse} and unusually large PL exponents, $\alpha\gg 6$, in particular for recent NLP models.
Our techniques, as implemented in the \emph{WeightWatcher} tool, can be used to identify when a pretrained DNN has problems that can not be detected simply by examining training/test accuracies.

### 2020-07-02 submission/introduction.tex
minor edits

Na\"{\i}vely---but in our experience commonly, among ML practitioners and ML theorists---if one does not have access to training or testing data, then one can say absolutely nothing about the quality of a ML model.
This may be true in worst-case theory, but models are used in practice, and there is a need for a \emph{practical theory} to guide that practice.
Moreover, if ML is to become an industrial process, then that process will become siloed: some groups will gather data, other groups will develop models, and other groups will use those models.
Users of models can not be expected to know the precise details of how models were built, the specifics of data that were used to train the model, what was the loss function or hyperparameter values, how precisely the model was regularized,~etc.

>>>>

Na\"{\i}vely---but in our experience commonly, among ML practitioners and ML theorists---if one does not have access to training or testing data, then one can say absolutely nothing about the quality of a ML model.
This may be true in worst-case theory, but models are used in practice, and there is a need for a \emph{practical theory} to guide that practice.
Moreover, if ML is to become an industrial process, then that process will become \nred{comparmentalized in ortder to scale}: some groups will gather data, other groups will develop models, and other groups will use those models.
Users of models can not be expected to know the precise details of how models were built, the specifics of data that were used to train the model, what was the loss function or hyperparameter values, how precisely the model was regularized,~etc.
